### 2022-07-15 draft.tex
Deploying to gh-pages from @ graciellehigino/ms_range_interactions@af47296c0554e5ff4be07181cc68d909ca2fb7f8 🚀

Finding a species in a certain location is like finding an encrypted
message that traveled through time. It carries the species' evolutionary
history, migration patterns, as well as any direct and indirect effects
generated by other species (some of which we may not even know exist).
Ecologists have been trying to decode this message with progressively
more powerful tools, from their field notes to highly complex
computational algorithms. However, to succeed in this challenge it is
important to have the right clues in hand. There are many ways we can be
misled by data - or the lack of it: taxonomic errors (e.g., due to
updates in the taxonomy of a species), geographic inaccuracy (e.g.,
approximate coordinates or lack of documentation about their accuracy),
or sampling biases (e.g.~data clustered near roads or research centers)
(Ladle and Hortal 2013; Hortal et al. 2015; Poisot et al. 2021). One way
to identify - and potentially fix - these errors is to combine many
different pieces of information about the occurrence of a species, so
agreements and mismatches can emerge. Although previous studies have
combined different types of occurrence data to measure the accuracy of
datasets (Hurlbert and Jetz 2007; Hurlbert and White 2005; Ficetola et
al. 2014), none have used different types of information so far (i.e.,
ecological characteristics other than geographical distribution). Here
we suggest jointly analysing species occurrence (range maps and point
occurrences) and ecological interactions to identify mismatches between
datasets.

@@

It has been demonstrated that the agreement between range maps and point
data varies geographically (Hurlbert and Jetz 2007; Hurlbert and White
2005; Ficetola et al. 2014). Adding ecological interaction data to this
comparison might help to elucidate where these (dis)agreements are more
likely to be true and which dataset better represent the actual
distribution of species. In this context, we elaborate a method that
allows us to refine distribution data (more precisely range maps) based
on interaction data, considering the basic assumption that predators can
only be present in regions where they are connected to at least one
herbivore - and thus indirectly connected to primary producers. We used
a Serengeti food web dataset (Baskerville et al. 2011) (which comprises
carnivores, herbivores, and plants from Tanzania) to demonstrate how a
mismatch between occurrence and interaction data can highlight
significant uncertainty areas in IUCN range maps. Finally, we add the
GBIF occurrence points for the Serengeti species to the investigation,
discuss the mechanisms that can lead to the lack of agreement between
data, and build from that a vision for the next steps, reinforcing the
importance of geographically explicit interaction data.

@@

We compiled IUCN range maps for the 32 species included in the metaweb
from the Spatial Data Download portal
(www.iucnredlist.org/resources/spatial-data-download), which we
rasterized at 10 arc-minute resolution (\textasciitilde19 km² at the
equator). We then combined interaction data from the metaweb and
cooccurrence data generated from species ranges to create networks for
each raster pixel. This generated a total of 84,244 pixel-level
networks. These networks describe potential predation, not actual
interactions: the former is derived information from the metaweb, and
the latter is contingent on the presence of herbivores.

@@

For each species in the dataset we collated point observation data from
GBIF (www.gbif.org), and condensed these data into pixels representing
presence or absence of the focal taxon. These data were used to validate
the range adjustments made based on species interactions (see the
previous section). To do so, we calculated the proportion of GBIF
presence pixels occurring within both the original and adjusted species
ranges. We then compared these proportions for the predators to verify
if the range adjustments removed locations with GBIF observations, hence
likely true habitats.

@@

Here we demonstrated how we can detect uncertainty in species
distribution data using ecological interactions. Knowing where
questionable occurrence data are can be crucial in ecological modelling
(Hortal 2008; Ladle and Hortal 2013), and accounting for these errors
can improve model outputs by diminishing the error propagation (Draper
1995). For instance, we believe this is a way to account for ecological
interactions in habitat suitability models without making the models
more complex, but by making sure (not assuming) that the input data -
the species occurrence - actually accounts for ecological interactions.
It is important to notice, however, that the quality and usefulness of
this method are highly correlated with the amount and quality of data
available about species' occurrences \textbf{and} interactions. With
this paper, we hope to add to the collective effort to decode the
encrypted message that is the occurrence of a species in space and time.
A promising avenue that adds to our method is the prediction of networks
and interactions in large scales (Strydom et al. 2021; Windsor et al.
2022), for they can add valuable information about ecological
interactions where they are missing. Additionally, in order to achieve a
robust modelling framework towards actual species distribution models we
should invest in efforts to collect and combine open data on species
occurrence and interactions (Windsor et al. 2022), especially because we
may be losing ecological interactions at least as fast as we are losing
species (Valiente-Banuet et al. 2015).

@@

@@

\leavevmode\hypertarget{ref-Dallas2020AbuNot}{}%
Dallas, Tad, Samuel Pironon, and Luca Santini. 2020. {``The
Abundant-Centre Is Not All That Abundant: A Comment to Osorio-Olvera Et
Al. 2020,''} 2020.02.27.968586.
\url{https://doi.org/10.1101/2020.02.27.968586}.

\leavevmode\hypertarget{ref-Daru2020GreToo}{}%
Daru, Barnabas H. 2020. {``GreenMaps: A Tool for Addressing the
Wallacean Shortfall in the Global Distribution of Plants.''}
\emph{bioRxiv}, 2020.02.21.960161.
\url{https://doi.org/10.1101/2020.02.21.960161}.

@@

\leavevmode\hypertarget{ref-Ficetola2014EvaRob}{}%
Ficetola, Gentile Francesco, Carlo Rondinini, Anna Bonardi, Vineet
Katariya, Emilio Padoa-Schioppa, and Ariadne Angulo. 2014. {``An
Evaluation of the Robustness of Global Amphibian Range Maps.''}
\emph{Journal of Biogeography} 41 (2): 211--21.
\url{https://doi.org/10.1111/jbi.12206}.

\leavevmode\hypertarget{ref-Fourcade2016ComSpe}{}%
Fourcade, Yoan. 2016. {``Comparing Species Distributions Modelled from
Occurrence Data and from Expert-Based Range Maps. Implication for
Predicting Range Shifts with Climate Change.''} \emph{Ecological
Informatics} 36: 8--14.
\url{https://doi.org/10.1016/j.ecoinf.2016.09.002}.

\leavevmode\hypertarget{ref-Fricke2022EffDef}{}%
Fricke, Evan C., Alejandro Ordonez, Haldre S. Rogers, and Jens-Christian
Svenning. 2022. {``The Effects of Defaunation on Plants' Capacity to
Track Climate Change.''} \emph{Science}.
\url{https://doi.org/10.1126/science.abk3510}.

@@

\leavevmode\hypertarget{ref-Poisot2020EnvBia}{}%
Poisot, Timothée, Gabriel Bergeron, Kevin Cazelles, Tad Dallas,
Dominique Gravel, Andrew Macdonald, Benjamin Mercier, Clément Violet,
and Steve Vissault. 2020. {``Environmental Biases in the Study of
Ecological Networks at the Planetary Scale.''} \emph{bioRxiv},
2020.01.27.921429. \url{https://doi.org/10.1101/2020.01.27.921429}.

@@

\leavevmode\hypertarget{ref-Ronquillo2020AssSpa}{}%
Ronquillo, Cristina, Fernanda Alves-Martins, Vicente Mazimpaka, Thadeu
Sobral-Souza, Bruno Vilela-Silva, Nagore G. Medina, and Joaquín Hortal.
2020. {``Assessing Spatial and Temporal Biases and Gaps in the Publicly
Available Distributional Information of Iberian Mosses.''}
\emph{Biodiversity Data Journal} 8: e53474.
\url{https://doi.org/10.3897/BDJ.8.e53474}.

@@

\leavevmode\hypertarget{ref-Scott2018RolHer}{}%
Scott, Abigail L., Paul H. York, Clare Duncan, Peter I. Macreadie, Rod
M. Connolly, Megan T. Ellis, Jessie C. Jarvis, Kristin I. Jinks, Helene
Marsh, and Michael A. Rasheed. 2018. {``The Role of Herbivory in
Structuring Tropical Seagrass Ecosystem Service Delivery.''}
\emph{Frontiers in Plant Science} 9: 127.
\url{https://doi.org/10.3389/fpls.2018.00127}.

@@

\leavevmode\hypertarget{ref-Valiente-Banuet2015SpeLos}{}%
Valiente-Banuet, Alfonso, Marcelo A. Aizen, Julio M. Alcántara, Juan
Arroyo, Andrea Cocucci, Mauro Galetti, María B. García, et al. 2015.
{``Beyond Species Loss: The Extinction of Ecological Interactions in a
Changing World.''} Edited by Marc Johnson. \emph{Functional Ecology} 29
(3): 299--307. \url{https://doi.org/10.1111/1365-2435.12356}.

@@

>>>>

Finding a species in a certain location is like finding an encrypted
message that traveled through time. It carries the species' evolutionary
history, migration patterns, as well as any direct and indirect effects
generated by other species (some of which we may not even know exist).
Ecologists have been trying to decode this message with progressively
more powerful tools, from their field notes to highly complex
computational algorithms. However, to succeed in this challenge it is
important to have the right clues in hand. There are many ways we can be
misled by data - or the lack of it: taxonomic errors (e.g., due to
updates in the taxonomy of a species), geographic inaccuracy (e.g.,
approximate coordinates or lack of documentation about their accuracy),
or sampling biases (e.g.~data clustered near roads or research centers)
(Ladle and Hortal 2013; Hortal et al. 2015; Poisot et al. 2021). One way
to identify - and potentially fix - these errors is to combine many
different pieces of information about the occurrence of a species, so
agreements and mismatches can emerge. Although previous studies have
combined different types of occurrence data to measure the accuracy of
datasets (Hurlbert and Jetz 2007; Hurlbert and White 2005;
\textbf{Ficetola2014EvaRob?}), none have used different types of
information so far (i.e., ecological characteristics other than
geographical distribution). Here we suggest jointly analysing species
occurrence (range maps and point occurrences) and ecological
interactions to identify mismatches between datasets.

@@

It has been demonstrated that the agreement between range maps and point
data varies geographically (Hurlbert and Jetz 2007; Hurlbert and White
2005; \textbf{Ficetola2014EvaRob?}). Adding ecological interaction data
to this comparison might help to elucidate where these (dis)agreements
are more likely to be true and which dataset better represent the actual
distribution of species. In this context, we elaborate a method that
allows us to refine distribution data (more precisely range maps) based
on interaction data, considering the basic assumption that predators can
only be present in regions where they are connected to at least one
herbivore - and thus indirectly connected to primary producers. We used
a Serengeti food web dataset (Baskerville et al. 2011) (which comprises
carnivores, herbivores, and plants from Tanzania) to demonstrate how a
mismatch between occurrence and interaction data can highlight
significant uncertainty areas in IUCN range maps. Finally, we add the
GBIF occurrence points for the Serengeti species to the investigation,
discuss the mechanisms that can lead to the lack of agreement between
data, and build from that a vision for the next steps, reinforcing the
importance of geographically explicit interaction data.

@@

We compiled IUCN range maps for the 32 species included in the metaweb
from the Spatial Data Download portal
(www.iucnredlist.org/resources/spatial-data-download), which we
rasterized at 10 arc-minute resolution (\textasciitilde18 km at the
equator). We restricted the rasters a spatial extent comprised between
latitudes 35°S and 40°N and longitudes 20°W and 55°E. We then combined
interaction data from the metaweb and cooccurrence data generated from
species ranges to create networks for each raster pixel. This generated
a total of 84,244 pixel-level networks. These networks describe
potential predation, not actual interactions: the former is derived
information from the metaweb, and the latter is contingent on the
presence of herbivores.

@@

For each species in the dataset we collated point observation data from
GBIF (www.gbif.org). We restricted our queries to the data with spatial
coordinates and which were inside the spatial extent of our rasters. We
did not use continental or date filters to retrieve as much data as
possible. However, a few observations were localized in the ocean near
latitude 0° and longitude 0°. We assumed these were errors and removed
all observations falling in the extent between latitudes 2°S and 2°N and
longitudes 2°W and 2°E to keep only mainland sites.

We then converted the occurrence data into raster format by determining
which pixels had a least one GBIF occurrence. This allowed us to remove
the effect of repeated sampling in some locations. These data were used
to validate the range adjustments made based on species interactions
(see beginning of Methods section). To do so, we calculated the
proportion of GBIF presence pixels occurring within both the original
IUCN species range and the adjusted one. We then compared these
proportions for the predators to verify if the range adjustments removed
locations with GBIF observations, hence likely true habitats.

\hypertarget{software}{%
\subsection{Software}\label{software}}

We performed all analyses using \emph{Julia} v1.7.2 (Bezanson et al.
2017). We used the packages \texttt{SimpleSDMLayers.jl} (Dansereau and
Poisot 2021) to manipulate the raster layers,
\texttt{EcologicalNetworks.jl} (Poisot et al. 2019) to construct and
manipulate the interaction networks, and \texttt{GBIF.jl} (Dansereau and
Poisot 2021) to retrieve the species occurrences from GBIF. We also used
\emph{GDAL} (GDAL/OGR contributors 2021) to rasterize the IUCN range
maps (initially available as shapefiles from the Spatial Data Download
portal). All the scripts required to reproduce the analyses are
available at {[}\textbf{link to Zenodo archive}{]}.

@@

Here we demonstrated how we can detect uncertainty in species
distribution data using ecological interactions. Knowing where
questionable occurrence data are can be crucial in ecological modelling
(Hortal 2008; Ladle and Hortal 2013), and accounting for these errors
can improve model outputs by diminishing the error propagation (Draper
1995). For instance, we believe this is a way to account for ecological
interactions in habitat suitability models without making the models
more complex, but by making sure (not assuming) that the input data -
the species occurrence - actually accounts for ecological interactions.
It is important to notice, however, that the quality and usefulness of
this method are highly correlated with the amount and quality of data
available about species' occurrences \textbf{and} interactions. With
this paper, we hope to add to the collective effort to decode the
encrypted message that is the occurrence of a species in space and time.
A promising avenue that adds to our method is the prediction of networks
and interactions in large scales (Strydom et al. 2021; Windsor et al.
2022), for they can add valuable information about ecological
interactions where they are missing. Additionally, in order to achieve a
robust modelling framework towards actual species distribution models we
should invest in efforts to collect and combine open data on species
occurrence and interactions (Windsor et al. 2022), especially because we
may be losing ecological interactions at least as fast as we are losing
species (\textbf{Valiente-Banuet2015SpeLos?}).

@@

\leavevmode\hypertarget{ref-Bezanson2017JulFre}{}%
Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B. Shah. 2017.
{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.

@@

\leavevmode\hypertarget{ref-Dallas2020AbuNot}{}%
Dallas, Tad, Samuel Pironon, and Luca Santini. 2020. {``The
Abundant-Centre Is Not All That Abundant: A Comment to Osorio-Olvera Et
Al. 2020,''} May, 2020.02.27.968586.
\url{https://doi.org/10.1101/2020.02.27.968586}.

\leavevmode\hypertarget{ref-Dansereau2021SimJl}{}%
Dansereau, Gabriel, and Timothée Poisot. 2021. {``SimpleSDMLayers.jl and
GBIF.jl: A Framework for Species Distribution Modeling in Julia.''}
\emph{Journal of Open Source Software} 6 (57): 2872.
\url{https://doi.org/10.21105/joss.02872}.

\leavevmode\hypertarget{ref-Dansereau2021SimJl}{}%
---------. 2021. {``SimpleSDMLayers.jl and GBIF.jl: A Framework for
Species Distribution Modeling in Julia.''} \emph{Journal of Open Source
Software} 6 (57): 2872. \url{https://doi.org/10.21105/joss.02872}.

\leavevmode\hypertarget{ref-Daru2020GreToo}{}%
Daru, Barnabas H. 2020. {``GreenMaps: A Tool for Addressing the
Wallacean Shortfall in the Global Distribution of Plants.''}
\emph{bioRxiv}, February, 2020.02.21.960161.
\url{https://doi.org/10.1101/2020.02.21.960161}.

@@

\leavevmode\hypertarget{ref-Fourcade2016ComSpe}{}%
Fourcade, Yoan. 2016. {``Comparing Species Distributions Modelled from
Occurrence Data and from Expert-Based Range Maps. Implication for
Predicting Range Shifts with Climate Change.''} \emph{Ecological
Informatics} 36 (November): 8--14.
\url{https://doi.org/10.1016/j.ecoinf.2016.09.002}.

\leavevmode\hypertarget{ref-Fricke2022EffDef}{}%
Fricke, Evan C., Alejandro Ordonez, Haldre S. Rogers, and Jens-Christian
Svenning. 2022. {``The Effects of Defaunation on Plants' Capacity to
Track Climate Change.''} \emph{Science}, January.
\url{https://doi.org/10.1126/science.abk3510}.

\leavevmode\hypertarget{ref-GDALux2fOGRcontributors2021GdaOgr}{}%
GDAL/OGR contributors. 2021. \emph{GDAL/OGR Geospatial Data Abstraction
Software Library}. Manual. Open Source Geospatial Foundation.

@@

\leavevmode\hypertarget{ref-Poisot2020EnvBia}{}%
Poisot, Timothée, Gabriel Bergeron, Kevin Cazelles, Tad Dallas,
Dominique Gravel, Andrew Macdonald, Benjamin Mercier, Clément Violet,
and Steve Vissault. 2020. {``Environmental Biases in the Study of
Ecological Networks at the Planetary Scale.''} \emph{bioRxiv}, January,
2020.01.27.921429. \url{https://doi.org/10.1101/2020.01.27.921429}.

\leavevmode\hypertarget{ref-Poisot2019EcoJl}{}%
Poisot, Timothée, Zachary Bélisle, Laura Hoebeke, Michiel Stock, and
Piotr Szefer. 2019. {``EcologicalNetworks.jl: Analysing Ecological
Networks of Species Interactions.''} \emph{Ecography} 42 (11): 1850--61.
\url{https://doi.org/10.1111/ecog.04310}.

@@

\leavevmode\hypertarget{ref-Ronquillo2020AssSpa}{}%
Ronquillo, Cristina, Fernanda Alves-Martins, Vicente Mazimpaka, Thadeu
Sobral-Souza, Bruno Vilela-Silva, Nagore G. Medina, and Joaquín Hortal.
2020. {``Assessing Spatial and Temporal Biases and Gaps in the Publicly
Available Distributional Information of Iberian Mosses.''}
\emph{Biodiversity Data Journal} 8 (September): e53474.
\url{https://doi.org/10.3897/BDJ.8.e53474}.

@@

\leavevmode\hypertarget{ref-Scott2018RolHer}{}%
Scott, Abigail L., Paul H. York, Clare Duncan, Peter I. Macreadie, Rod
M. Connolly, Megan T. Ellis, Jessie C. Jarvis, Kristin I. Jinks, Helene
Marsh, and Michael A. Rasheed. 2018. {``The Role of Herbivory in
Structuring Tropical Seagrass Ecosystem Service Delivery.''}
\emph{Frontiers in Plant Science} 9 (February): 127.
\url{https://doi.org/10.3389/fpls.2018.00127}.

@@

@@

\leavevmode\hypertarget{ref-Windsor2022UsiEco}{}%
---------. 2022. {``Using Ecological Networks to Answer Questions in
Global Biogeography and Ecology.''} \emph{Journal of Biogeography} n/a
(n/a). \url{https://doi.org/10.1111/jbi.14447}.
